---
title: "P8105 Homework 2"
author: "Quinton Neville"
date: "Due October 4th, 2018"
header-includes: 
  \usepackage{graphicx}
  \usepackage{float}
output:
   github_document 
---

```{r setup, echo = FALSE, warning = FALSE, message = FALSE}
#Load necessary packages
library(tidyverse)
library(readxl)
library(readr)
library(p8105.datasets)
```

This problem focuses on NYC Transit data; in particular, this CSV file contains information related to each entrance and exit for each subway station in NYC.

Read and clean the data; retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (the ifelse or recode function may be useful).

Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?  

Answer the following questions using these data:  

How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St A/B/C/D; 125st 1; 125st 4/5); the distinct function may be useful here.  

How many stations are ADA compliant?
What proportion of station entrances / exits without vending allow entrance?
Reformat data so that route number and route name are distinct variables. How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?

#Problem 1
```{r}
nyc.transit.df <- read_csv(file = "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
                           col_types = cols(
                               Route8 = col_character(),
                               Route9 = col_character(),
                               Route10 = col_character(),
                               Route11 = col_character()
                           )) %>%
  janitor::clean_names() %>%
  select(line:route11, entry, vending, entrance_type, ada) %>%
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))

#Dimensions
dim.nyc.transit <- dim(nyc.transit.df)   #Vector of [1] Rows, [2] Columns

#Count Distinct Stations by Line and Name
collapse.route <- function(x){
  paste0(na.omit(x), collapse = "/")
}

distinct.route.df <- nyc.transit.df %>%
  mutate(route = apply(select(., route1:route11), 1, collapse.route)) %>%
  select(line, station_name, route, everything(), -c(route1:route11))

a <- distinct.route.df %>% distinct(., line, station_name) #465

b <- distinct.route.df %>% distinct(., line, station_name, route) #467

distinct.stations <-  distinct.route.df %>% distinct(., line, station_name) %>% nrow()

#Count distinct stations ADA compiance
distinct.stations.df <- nyc.transit.df %>%
  distinct(., line, station_name, ada)

```

