---
title: "P8105 Homework 2"
author: "Quinton Neville"
date: "Due October 4th, 2018"
header-includes: 
  \usepackage{graphicx}
  \usepackage{float}
output:
   github_document 
---

```{r setup, echo = FALSE, warning = FALSE, message = FALSE}
#Load necessary packages
library(tidyverse)
library(readxl)
library(readr)
library(p8105.datasets)
```

This problem focuses on NYC Transit data; in particular, this CSV file contains information related to each entrance and exit for each subway station in NYC.

Read and clean the data; retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (the ifelse or recode function may be useful).

Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?  

Answer the following questions using these data:  

How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St A/B/C/D; 125st 1; 125st 4/5); the distinct function may be useful here.  

How many stations are ADA compliant?
What proportion of station entrances / exits without vending allow entrance?
Reformat data so that route number and route name are distinct variables. How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?

#Problem 1
```{r}
nyc.transit.df <- read_csv(file = "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
                           col_types = cols(
                               Route8 = col_character(),
                               Route9 = col_character(),
                               Route10 = col_character(),
                               Route11 = col_character()
                           )) %>%
  janitor::clean_names() %>%
  select(line:route11, entry, vending, entrance_type, ada) %>%
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))

#Dimensions
dim.nyc.transit <- dim(nyc.transit.df)   #Vector of [1] Rows, [2] Columns

#Count Distinct Stations by Line and Name
collapse.route <- function(x){
  paste0(na.omit(x), collapse = "/")
}

distinct.route.df <- nyc.transit.df %>%
  mutate(route = apply(select(., route1:route11), 1, collapse.route)) %>%
  select(line, station_name, route, everything(), -c(route1:route11))

a <- distinct.route.df %>% distinct(., line, station_name) #465 Correct, if #109 is invalid. Assuming #1628 miscoded

b <- distinct.route.df %>% distinct(., line, station_name, route) #467
  
#OBS 109 in original data has an extra train serviced at one of the entrances (6thAV & 47-50th Sts Rockefeller Center, extra 7 train)
#OBS 1628 in original data has a miss-coded e instead of E at one entrance (Queens & Forest Hills, route1 = e)

distinct.stations <-  distinct.route.df %>% distinct(., line, station_name) %>% nrow()

#Count distinct stations ADA compiance
distinct.stations.df <- distinct.route.df %>%
  distinct(., line, station_name, route, ada)

distinct.ada.comp <- sum(distinct.stations.df$ada, na.rm = TRUE)
distinct.ada.prop <- mean(distinct.stations.df$ada, na.rm = TRUE)

#Proportion of entries/exits with no vending that allow entrance
no.vend.entry <- distinct.route.df %>%
  filter(vending == "NO") %>%
  select(entry) %>%
  apply(., 2, mean)

#Pulling those which service the A train and counting how many are ADA compliant
distinct.stations.df %>%
  mutate(service_a = grepl("A", route)) %>%
  filter(service_a == TRUE) %>%
  select(ada) %>%
  apply(., 2, sum)

#Note add cleaning steps from the original data frame in each problem, i.e. pipe the whole thing frome original
```

Problem2
This problem uses the Mr. Trash Wheel dataset, available as an Excel file on the course website. Please use the  HealthyHarborWaterWheelTotals2017-9-26.xlsx version.

Read and clean the Mr. Trash Wheel sheet:

specify the sheet in the Excel file and to omit columns containing notes (using the range argument and cell_cols() function)
use reasonable variable names
omit rows that do not include dumpster-specific data
rounds the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)
Read and clean precipitation data for 2016 and 2017. For each, omit rows without precipitation data and add a variable year. Next, combine datasets and convert month to a character variable (the variable month.name is built into R and should be useful).

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in both resulting datasets, and give examples of key variables. For available data, what was the total precipitation in 2017? What was the median number of sports balls in a dumpster in 2016?

#Problem 2
```{r}
###Mr. Trashwheel###
#Read and Clean Mr. Trash Wheel Data
trash.wheel.df <- read_xlsx("./data/HealthyHarborWaterWheelTotals2017-9-26.xlsx",
                            sheet = "Mr. Trash Wheel", range = cell_cols("A:N")) %>%  #Select only variable columns
                  janitor::clean_names() %>%                                          #Standardize variable names
                  mutate(total = grepl("Total", month),                               #Create logical for "Total", for later filtering
                         sports_balls = as.integer(sports_balls),                     #Round sports_balls to nearet integer
                         year = ifelse(is.na(year), "2017", year)) %>%                #Impute missing year to match date variable
                  filter(total == FALSE) %>%                                          #Omit all rows with non-dumpster spec. data
                  select(-total) %>%                                                  #Remove logical total variable
                  na.omit()                                                           #Omit the remaining NA rows

#Key Stats

#Dimension
dim.trash.wheel <- dim(trash.wheel.df) #c(Rows, Cols)

#Create Tibble of Summary stats to access easily inline
total.trash.df <- trash.wheel.df %>% 
              select(weight_tons:homes_powered) %>%           #Taking all the cont. variables, calculating sum/totals, store as tibble
              apply(., 2, sum) %>% 
              t() %>%
              as.tibble()
summary.trash.df <- trash.wheel.df %>% 
              select(weight_tons:homes_powered) %>%           #Taking all the cont. variables, calc. summary stats, store as tibble
              apply(., 2, summary) %>%                        #Row bind with total
              as.tibble() %>%                                 #Add 'statistic' variable to access easiliy by filter inline
              bind_rows(., total.trash.df) %>%
              mutate(statistic =  c("Min", "1Q", "Median", "Mean", "3Q", "Max", "Sum")) %>%
              select(statistic, everything())

#Calculate Median Sports Balls in 2016
median.sports.balls <- trash.wheel.df %>%
                       filter(year == "2016") %>%
                       select(sports_balls) %>%
                       apply(., 2, median)

###2016-2017 Precipitation###
#Read 2016-2017 Precipitation Data
precip.16.df <- read_xlsx("./data/HealthyHarborWaterWheelTotals2017-9-26.xlsx",
                            sheet = "2016 Precipitation", range = "A2:B14")

precip.17.df <- read_xlsx("./data/HealthyHarborWaterWheelTotals2017-9-26.xlsx",
                            sheet = "2017 Precipitation", range = "A2:B14")

#Clean 2016-2017 Precipitation Data
precip.years <- c("2016", "2017")  #Years of Precipitation observation

precip.16.17.df <- bind_rows(precip.16.df, precip.17.df) %>%                             #Row bind 16-17 data
                   janitor::clean_names() %>%                                            #Clean names
                   mutate(month = rep(month.name, times = length(precip.years)),         #Change month variable to char
                          year = rep(precip.years, each = length(month.name))) %>%       #Create variable for year of correct length
                   filter(!is.na(total))                                                 #Omit obs. w/no total reported

#Key Stats
dim.precip.16.17 <- dim(precip.16.17.df)

#Total for both years
total.precip.df <- precip.16.17.df %>% 
                   select(total) %>%              #Taking continuous variable Total
                   apply(., 2, sum) %>%           #Counting the total over both years
                   t() %>% as.tibble()

#Summary monthly rainfall for both years & total in one tibble
summary.precip.df <- precip.16.17.df %>%
                 select(total) %>%                #Taking cont. variable Total
                 apply(., 2, summary) %>%         #Summarazing it over both years
                 as.tibble() %>%                  #Binding with total for easy inline access
                 bind_rows(., total.precip.df) %>%
                 mutate(statistic =  c("Min", "1Q", "Median", "Mean", "3Q", "Max", "Sum")) %>%
                 select(statistic, everything())

#Total Precipitation in 2017
total.precip.17 <- precip.16.17.df %>%
                   filter(year == 2017) %>%
                   select(total) %>%
                   sum()
```

